Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-02 sections/eval.tex
Update on Overleaf.

We now describe how to define a few example functions from the API in this language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$. For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}
Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops. In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels. However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.

\subsubsection{Standardizing \& Heuristic Optimization}
Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves the above questions about loop ordering and output formatting. This standard form has a few syntactic requirements, but the semantically important requirements are 1) all inputs (i.e. tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator. The former allows the interpreter to identify the loop order for each kernel. The second determines the output format for each intermediate. The last one guarantees that the innermost expression can be computed as a single kernel.

During the standardization process, a concordization pass is performed which examines each query in order and selects a loop order based on a heuristic which loops over intersecting variables first. Then, each query is examined again and a transposition query is inserted for each input which doesn't match the loop order and then the input is replaced with an alias. At this point, the first semantic requirement is satisfied. Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required. The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.

>>>>

We now describe how to define a few example functions from the API in this language. Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$. For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
\begin{align*}
&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
\end{align*}

\subsubsection{Standardizing \& Heuristic Optimization}
Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting. The semantically important requirements of this standard form are 1) all inputs (i.e. tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator. The former allows the interpreter to identify the loop order for each kernel. The second determines the output format for each intermediate. The last one guarantees that the innermost expression can be computed as a single kernel.

During the standardization process, a concordization pass is performed which examines each query, heuristically selects a loop order, and transposes inputs which don't match that order. Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required. The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.
